---
title: "STORMDATA"
author: "LUIS V"
date: "3/10/2021"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE , warning = FALSE)
```

# STORMDATA Project Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## DATA Processing

### Loading and preprocessing the data
```{r readdata}
library(dplyr)
library(tidyr)
library(R.utils)
library(lubridate)
library(ggplot2)

setwd("C:/Users/ventd/OneDrive/Escritorio/Coursera/Reproducible Research/RRProject2")

filename <- "Dataset.csv.bz2"

##checking if the file already exists, if it doesn't it will download it
if (!file.exists(filename)){
        fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
        download.file(fileURL, filename, method="curl")
} 
if (!file.exists("Dataset")) { 
        xdata <- read.csv(bzfile(filename)) 
}
##Reading data
xdata <- read.csv(bzfile(filename))
```

### Calculating and analizing data 

```{r calcdata}
dim(xdata)
str(xdata)
```

### Getting relevant variants for this project which contains the following questions:

        1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful                 with respect to population health?
        2. Across the United States, which types of events have the greatest economic consequences?

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

```{r newdata}
#Creating the table with variables we are going to use
pdata <- select(xdata, BGN_DATE, STATE, EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP, FATALITIES, INJURIES)

```

Checking if there are any NA's in the data
```{r}
sum(is.na(pdata))
```

# Start to Cleaning Data

Format the BGN_DATE variable as a date
Latest years are the ones with more relevant information, 48 event since 1996
```{r}
pdata$BGN_DATE <- as.Date(pdata$BGN_DATE, "%m/%d/%Y")
pdata$YEAR <- year(pdata$BGN_DATE)
pdata <- filter(pdata, YEAR >= 1996)
```

Only use events with either health impact or economic damage
```{r}
pdata <- filter(pdata, PROPDMG > 0 | CROPDMG > 0 | FATALITIES > 0 | INJURIES > 0)
```

### Getting and cleaning variables to get the correct information for the economic consequences graphic
```{r}
table(pdata$PROPDMGEXP)
table(pdata$CROPDMGEXP)
pdata$PROPDMGEXP <- toupper(pdata$PROPDMGEXP)
pdata$CROPDMGEXP <- toupper(pdata$CROPDMGEXP)
```

### Converting, calculating, and adding new variables needed to get the graphics.
```{r}
pdata <- pdata %>% 
        mutate(CROPDMGFACTOR = case_when(CROPDMGEXP == "" ~ 10^0 * CROPDMG,
                                         CROPDMGEXP == "?" ~ 10^0 * CROPDMG,
                                         CROPDMGEXP == "0" ~ 10^0 * CROPDMG,
                                         CROPDMGEXP == "2" ~ 10^2 * CROPDMG,
                                         CROPDMGEXP == "K" ~ 10^3 * CROPDMG,
                                         CROPDMGEXP == "M" ~ 10^6 * CROPDMG,
                                         CROPDMGEXP == "B" ~ 10^9 * CROPDMG)) %>%
        mutate(PROPDMGFACTOR = case_when(PROPDMGEXP == "" ~ 10^0 * PROPDMG,
                                         PROPDMGEXP == "-" ~ 10^0 * PROPDMG,
                                         PROPDMGEXP == "?" ~ 10^0 * PROPDMG,
                                         PROPDMGEXP == "+" ~ 10^0 * PROPDMG,
                                         PROPDMGEXP == "0" ~ 10^0 * PROPDMG,
                                         PROPDMGEXP == "1" ~ 10^1 * PROPDMG,
                                         PROPDMGEXP == "2" ~ 10^2 * PROPDMG,
                                         PROPDMGEXP == "3" ~ 10^3 * PROPDMG,
                                         PROPDMGEXP == "4" ~ 10^4 * PROPDMG,
                                         PROPDMGEXP == "5" ~ 10^5 * PROPDMG,
                                         PROPDMGEXP == "6" ~ 10^6 * PROPDMG,
                                         PROPDMGEXP == "7" ~ 10^7 * PROPDMG,
                                         PROPDMGEXP == "8" ~ 10^8 * PROPDMG,
                                         PROPDMGEXP == "H" ~ 10^2 * PROPDMG,
                                         PROPDMGEXP == "K" ~ 10^3 * PROPDMG,
                                         PROPDMGEXP == "M" ~ 10^6 * PROPDMG,
                                         PROPDMGEXP == "B" ~ 10^9 * PROPDMG,)) %>%
        mutate(SUMDMG = PROPDMGFACTOR+CROPDMGFACTOR) %>%
        mutate(SUMFATINJ = FATALITIES + INJURIES)
```

Check if there is any NA in the new 2 columns
```{r}
sum(is.na(pdata))
```

### Summarizing data after cleaning 
```{r}
sumpdata <- pdata %>%
        group_by(EVTYPE) %>%
        summarize(SUMFATALITIES = sum(FATALITIES),
                  SUMINJURIES = sum(INJURIES),
                  TOTALFATINJ = sum(SUMFATINJ),
                  SUMPROPDMG = sum(PROPDMGFACTOR),
                  SUMCROPDMG = sum(CROPDMGFACTOR),
                  TOTALDMG = sum(SUMDMG))
head(sumpdata)
```
# Results

Harmful Impact with respect population Results
```{r}
Impact <- arrange(sumpdata, desc(TOTALFATINJ))
ImpactData <- head(Impact,10)

ImpactData
```

Economic Impact Results
```{r}
Economic <- arrange(sumpdata, desc(TOTALDMG))
EconomicData <- head(Economic,10)

EconomicData
```

## Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
```{r}
ImpactData$EVTYPE <- with(ImpactData, reorder(EVTYPE, -TOTALFATINJ))
ImpactDataS <- ImpactData %>%
        gather(key = "Type", value = "TOTALIMPACT", c("SUMFATALITIES", "SUMINJURIES")) %>%
        select(EVTYPE, Type, TOTALIMPACT)
ImpactDataS$Type[ImpactDataS$Type %in% c("SUMFATALITIES")] <- "Fatalities"
ImpactDataS$Type[ImpactDataS$Type %in% c("SUMINJURIES")] <- "Injuries"

g1 <- ggplot(ImpactDataS, aes(x = EVTYPE, y = TOTALIMPACT, fill = Type)) +
        geom_bar(stat = "identity", position = "stack") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
        xlab("Event Type") +
        ylab("Total Health Impact") +
        ggtitle("Events with Most Health Impact") +
        theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
g1
```


## Across the United States, which types of events have the greatest economic consequences?
```{r}
EconomicData$EVTYPE <- with(EconomicData, reorder(EVTYPE, -TOTALDMG))
EconomicDataS <- EconomicData %>%
        gather(key = "Type", value = "TOTALDAMAGE", c("SUMPROPDMG", "SUMCROPDMG")) %>%
        select(EVTYPE, Type, TOTALDAMAGE)
EconomicDataS$Type[EconomicDataS$Type %in% c("SUMPROPDMG")] <- "Property damage"
EconomicDataS$Type[EconomicDataS$Type %in% c("SUMCROPDMG")] <- "Crop damage"

g2 <- ggplot(EconomicDataS, aes(x = EVTYPE, y = TOTALDAMAGE, fill = Type)) +
        geom_bar(stat = "identity", position = "stack") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
        xlab("Event Type") +
        ylab("Total Economic Impact") +
        ggtitle("Events with Most Economic Impact") +
        theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
g2

```

## Conclusion

Per the graphics we can conclude what events are the most dangerous in economic 
and health, if we can get a better way to anticipate and evacuate people this 
can help in reduce the danger for people, and getting better prepare for flood 
in materials can help to reduce the economic impact.